A Plasma Torus Around a Young Low-Mass Star

### 2025-04-23 paper/ms.tex
update

\received{--}
\revised{--}
\accepted{--}
\shorttitle{M Dwarf Plasma Torus}

@@

\begin{abstract}
A small fraction of red dwarfs younger than $\sim$100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star.  However,
the composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces.  Here, we present
time-series spectroscopy and photometry of a rapidly-rotating
($P$=3.9\,hr) CPV, TIC\,141146667.  The spectra show sinusoidal
time-varying Balmer emission at twice to four times the star's
equatorial velocity, providing direct evidence for cool
($\lesssim$10$^4$\,K) plasma clumps trapped in corotation around a CPV.
These data support the idea that young, rapidly-rotating M dwarfs can
sustain warped plasma tori, similar to massive magnetic stars.  The
origin of the optically thick material in these tori, and the issue
of whether this material contains a significant amount of dust,
remain unclear.
\end{abstract}

\keywords{Circumstellar matter (241), Stellar magnetic fields (1610),
Stellar rotation (1629) Periodic variable stars (1213), Weak-line T
Tauri stars (1795) }

@@

One glaring example of our current ignorance is the complex periodic
variables.  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous systems
have now been found by K2 and TESS
\citep{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Stauffer2021,Popinchalk2023,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \citep{Stauffer2017,Bouma2024}.  However,
$\approx$3\% of stars a few million years old show this complex
behavior, an observed fraction which decreases to
$\approx$0.3\% by $\approx$110\,Myr \citep{Rebull2020}.  CPVs can and
have been confused for transiting exoplanets
\citep{vanEyken2012,Johns-Krull2016,Bouma2020}.

@@

We identified TIC\,141146667 in previous work \citep{Bouma2024} by
searching TESS two-minute data from 2018-2022 for highly
structured, periodic light curves \citep{Ricker2015}.  We selected the
star for spectroscopy because its brightness and rapid rotation enabled
an efficient search for variability in its line profiles.
As an apparently single pre-main sequence M dwarf, its properties are
typical for the CPV population (see Table~\ref{tab:params} and
Appendix~\ref{sec:stparams}).

TESS observed TIC\,141146667 ($T$=13.3) for six non-contiguous months
spanning 2019-2024.  TESS acquired these observations at two-minute
cadence during Sectors 41, 48 (TESS DDT039, PI: Kunimoto), and 75
(TESS Program G06030, PI: Bouma).   The 30-minute data during Sectors
14, 15, and 21 probably smeared sharp features over the $<$4\,hour
period (see \citealt{Gunther2022}).  The movie in
Figure~\ref{fig:lc} shows the two-minute data: 
like other CPVs, TIC\,141146667 maintains a fixed period over
timescales of years while its detailed photometric morphology evolves.
The nearest known star, TIC\,141146666 ($T$=14.5), is 25$''$ from
TIC\,141146667 and is photometrically stable in the TESS images;
crowding is not a concern.

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.935\textwidth]{figures/sf1.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Evolution of TIC\,141146667 during the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps caused by light from Earth (BTJD 3356-3358.5) and Moon
  (BTJD 3359.5-3360.5) are visible.  Raw two minute
  data are in gray; black time-averages to ten minute sampling.  {\bf
  b-c,} Folded TESS light curve before and after spectroscopy.  Black
  now phase-averages to 100 points per
  cycle.  During BTJD 3352-3356, a state switch occurred near BTJD
  3353, and the dip at $\phi$$\approx$0.8 disappeared.  While the
  large flux decrement was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

While the data gap is unfortunate, Figure~\ref{fig:fulllc} shows
that a large flux decrement spanned roughly half of each
cycle both before and after the HIRES data were acquired.  From BTJD
3352-3356, this dip had two sharp local minima;  the local minimum
at $\phi$$\approx$0.8 disappeared following the flare at BTJD 3353,
yielding a dip more closely resembling an asymmetric ``V'' than a ``W''.
Similar CPV state changes have been previously noted 
\citep{Stauffer2017,Bouma2024}.  The photometric shape therefore
evolved during the twelve cycles spanning the 3356-3358 gap, since
the average shape from 3358-3363 more closely resembles the initial
``W''.  Nonetheless, the general photometric morphology---a small
brightening over 45\% of the period, followed by a complex flux dip
spanning 55\% of the period---is similar before and after the data
gap.

Figure~\ref{fig:spec} compiles the TESS and HIRES data from February
2024.  The HIRES spectra show emission in H$\alpha$ beyond the star's
equatorial velocity, $v_{\rm eq}$, of 130\,\kms.  There are at least
two distinct emission components, separated by half a cycle in phase.
The inner component at lower velocities has clearer sinusoidal
behaviour in time and is double-peaked, with peak semi-amplitudes of
2.07\,$v_{\rm eq}$ and 2.88\,$v_{\rm eq}$, measured following the
procedure in Appendix~\ref{subsec:halpha}.  There is significant
non-periodic variability in the emissivity of this double-peaked
component: the flux excess begins with an amplitude 70\% that of the
continuum, and diminishes to 10\% by the end.  The higher-velocity
component 180$^\circ$ opposite in phase is detected only from
$\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this outer component appears
connected to the star in velocity space.  While its peak
semi-amplitude of 3.88\,$v_{\rm eq}$ is achieved at both $\phi$=0.25
and 0.75, its amplitude similarly decreases from a 60\% excess over
the continuum at the beginning of the observation sequence to a 10\%
excess by its end.  
From the same procedure in Appendix~\ref{subsec:halpha},
we measured
the period for all three spectroscopic
emission components to be consistent with the photometric \periodhr\ hour
period, to within two minutes for the lower-velocity component.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and velocities, this material's motion is
not controlled by gravitational attraction to the star; it is more
readily explained as plasma being dragged along
with a rigidly rotating stellar magnetic field.  The velocity
semi-amplitude of the sinusoids gives the mean distance of each clump
from the stellar surface: 2.07\,$R_\star$ and 2.88\,$R_\star$ for the
inner clumps, and 3.88\,$R_\star$ for the outer clump.   These clumps
transit in front of the star when passing from negative to positive
velocity.  The transits of the two inner clumps last $\approx$22\% of
each cycle, from $\phi$=-0.1 to $\phi$=+0.12 (see
Appendix~\ref{subsec:halpha}).  This spectroscopic transit coincides with
the latter half of the complex eclipse feature in the TESS data.

The H$\alpha$ line core is more complex.  At $|\Delta v / v_{\rm
eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material might also modulate the line
profile.  Figure~\ref{fig:spec}e suggests line core variability caused
by both bright and dark regions on the star's surface, superposed with
smaller-amplitude variability from the transiting circumstellar
material.  For instance, from $\phi$=0-0.3, the double-peaked emission
feature is visible when viewed both on and off-limb; this feature is
circumstellar in origin.  However, the large emission feature that
crosses the star from $\phi$=0.4-0.9 emits at an amplitude greater
than that observed from the circumstellar components, and it crosses
the stellar velocity surface at a speed that suggests it instead
originates from a chromospherically bright region on the star's
surface.  Similarly, from $\phi$=0.6-1.15 a 20\% deep absorption
feature slowly crosses the H$\alpha$ line profile.  This feature
suggests either a chromospherically dark region (e.g.~a starspot
group) crossing the star's surface, or an azimuthally extended
absorptive component to the circumstellar material which is not
visible in emission.  Although the origin of the other bright and dark
streaks passing across the line core are similarly ambiguous, a final
exercise to quantify the behavior of the line core is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

@@

\caption{Selected system parameters for TIC\,141146667}

@@

@@

\subsection{Corotating Plasma Clumps}

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they demonstrate that corotating circumstellar plasma
clumps exist around at least one such star.  While starspots do
contribute smooth signals to the photometric variability, a
``starspot-only'' scenario to explain the complexity of the CPVs
\citep{Koen2021} is incompatible with our observations.  Scenarios in
which the circumstellar material is made only of dust are similarly
ruled out.  While starspots and dust may both be present, our HIRES
spectra require plasma with a significant population of hydrogen atoms
in the $n$=3 excited state, with no evidence for significant
higher-order excitations (see Appendix~\ref{subsec:specvar}).  Most of
this H$\alpha$ emission originates in ``clumps'' with size similar to
that of the star; radial ``spokes'' or azimuthally-extended ``arcs''
for the emitting material are ruled out by the observed H$\alpha$
velocity dispersion (see Appendix~\ref{subsec:halpha}).  The observed
H$\alpha$ luminosity suggests characteristic densities and masses for
the gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
Appendix~\ref{subsec:gas}).  Dust, if present, is independently
constrained to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} <
10^{17}\,{\rm g}$ (see Appendix~\ref{subsec:dust}).

%TODO: improve heading?
\subsection{Plasma Tori Could Explain Complex Periodic Variables}

While the H$\alpha$ emission appears consistent with clumps, the
photometric width of the complex flux dips in TIC\,141146667
specifically, and CPVs more broadly \citep[Figure~4 of][]{Bouma2024},
raise the possibility that azimuthally-distributed material may be the
norm for these objects.  In other words, a torus may be a more
accurate picture than a clump.

\citet{Townsend2005} outline the physics of how such structures can
exist: for stars with magnetospheric radii $R_{\rm m}$ that exceed
their corotation radii $R_{\rm c}$, the effective potential
experienced by a charged particle can have a local minimum outside
$R_{\rm c}$, enabling material to accumulate in a
centrifugally-supported magnetosphere
\citep{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, \citet{Townsend2008} synthesized light curves for plasma
trapped in rigidly rotating magnetospheres of rapidly rotating stars;
and found that assuming a dipole magnetic geometry, one can obtain
W-shaped eclipses similar to those seen for TIC\,141146667.  See in
particular the
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}
associated with their work; for a $\omega/\omega_{\rm c}$$\approx$0.5,
edge-on system like TIC\,141146667, the eclipses have the correct
approximate shape, and the H$\alpha$ emission can similarly exhibit
double-peaked behavior.

\subsection{Origin of Circumstellar Material and Astrophysical Analogs}

The circumstellar material could originate either from the star or
from an external source.  Plausible external sources include an
undetected old disk, comets, or a close-in exoplanet.  This latter
scenario would make CPVs extrasolar analogs of the Jupiter-Io plasma
torus (e.g.~\citealt{Bagenal1981}), although with a very different
composition.

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
which accumulate into warped plasma tori
\citep{Townsend2005,Townsend2008}.  These tori tend to have dense
antipodal accumulations of plasma sculpted by tilted-dipole magnetic
fields, and the transits of these clumps produce broadband photometric
variability through bound-free scattering \citep{Townsend2005} and
Thomson scattering \citep{Berry2022}.  The plasma accumulates at
antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic planes (see Equation~22 of
\citealt{Townsend2005}).  For $\sigma$~Ori~E and almost all of its
analogs, the result is ``simple'' light curves that resemble those of
eclipsing binaries, and time-dynamic H$\alpha$ spectra similar to
those in Figure~\ref{fig:spec} \citep{Townsend2005,Townsend2008}.  The
two known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \citep{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\citep{Kochukhov2011,Shultz2018}.  There are two implications: first,
the photometric complexity of CPVs may be a direct consequence of
magnetic fields with highly multipolar contributions.  Second, despite
this complexity, the observation (Figure~\ref{fig:spec}) of two
emission clumps separated 180$^\circ$ in phase is consistent with the
expected emission from a warped plasma torus.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit 
sinusoidal time-varying Balmer emission
\citep{Donati2000,Townsend2005,Dunstone2006,Skelly2008}, similar to
that Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \citep{Bouma2024}.   One interpretation for the
spectroscopic variability of these stars, and that of the analogous transient
absorption features
\citep{CollierCameron1989,CollierCameron1992,Cang2020}, comes from an
analogy to quiescent solar prominences, cool condensations of plasma
in the solar corona that can last days to weeks
\citep{VialEngvold2015}.  In the case of the Sun, these condensations
fall back to the photosphere because gravity is stronger than any
magnetic tension or centrifugal force capable of sustaining them.  
However it has been understood at least since work by
\citet{Donati2000} that these ``prominence systems'' can be
longitudinally extended, forming a trapped ring of plasma.
While the question of why these previously known spectroscopic
variables do not show complex photometric variability remains open,
a possible explanation is that they are not in the required transiting
geometry.
What we have in the case of TIC\,141146667 is a system that finally
exhibits both sets of phenomena.

@@

Pressing issues for future work include determining the composition
and origin of the circumstellar material, understanding the exact role
of the stellar magnetic field, and exploring the implied space weather
experienced by the close-in rocky exoplanets that, statistically
\citep{Dressing2015}, are likely to be present in most CPV systems.

@@

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
magnetohydrodynamic modeling to go beyond previous work
\citep{Townsend2005,Townsend2008,Krticka2022} and to explore what field
topologies might explain the observed light curve morphologies
\citep{Bouma2024}.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \citep{Brown2020}; such fields
would yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent constraint on magnetic field topology may also come from
radio observations, which have shown \citep{Kaur2024} that CPVs emit
variable radio signals, including persistent and polarized components.
Detecting radio emission produced by an electron cyclotron maser
instability (e.g.~\citealt{Callingham2021}) in particular would provide a
measurement of the field strength at the site of the emitting region,
further clarifying the magnetosphere's structure.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\citep{Dressing2015}.  However, a few percent of young M dwarfs show
the CPV phenomenon \citep{Rebull2020}, and our data show that some of
the complex photometric features occur when clumps of circumstellar
material transit the star.  The suggested geometric correction implies
that an appreciable minority (10-30\%) of young M dwarfs -- the
rapidly rotating ones with centrifugal magnetospheres -- host
circumstellar environments similar to the CPVs.  Future studies that
combine spectroscopic, polarimetric, and multi-wavelength
observations, along with magnetohydrodynamic modeling, will be key to
understanding the complex magnetospheric environments of these young
stars.

>>>>

\received{--}
\revised{--}
\accepted{--}
\shorttitle{CPV Plasma Torus}

@@

\begin{abstract}
  A small fraction of red dwarfs younger than $\sim$100 million years
  show structured, periodic optical light curves suggestive of
  transiting opaque material that corotates with the star.  However, the
  composition, origin, and even the existence of this material are
  uncertain. The main alternative hypothesis is that these complex
  periodic variables (CPVs) are explained by complex distributions of
  bright or dark regions on the stellar surfaces.  Here, we present
  time-series spectroscopy and photometry of a rapidly-rotating
  ($P$=3.9\,hr) CPV, TIC\,141146667.  The spectra show sinusoidal
  time-varying H$\alpha$ emission at twice to four times the star's
  equatorial velocity, providing direct evidence for cool
  ($\lesssim$10$^4$\,K) plasma clumps trapped in corotation around a
  CPV.  These data support the idea that young, rapidly-rotating M
  dwarfs can sustain warped plasma tori, similar to massive magnetic
  stars.  It remains unclear whether these structures originate
  intrinsically from the star or are fed by external sources. Likewise,
  the mechanism by which their optically thick clumps arise is still
  unknown.  Rough estimates suggest $\gtrsim$10\% of M dwarfs host
  similar structures during their early lives.
\end{abstract}

\keywords{Circumstellar matter (241), Stellar magnetic fields (1610),
Stellar rotation (1629) Periodic variable stars (1213), Weak-line T
Tauri stars (1795)}

@@

One example of our current ignorance is the complex periodic
variables.  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous systems
have now been found by K2 and TESS
\citep{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Stauffer2021,Popinchalk2023,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \citep{Stauffer2017,Bouma2024}.  However,
$\approx$3\% of stars a few million years old show this complex
behavior, an observed fraction which decreases to
$\approx$0.3\% by $\approx$110\,Myr \citep{Rebull2020}.  CPVs can and
have been confused for transiting exoplanets
\citep{vanEyken2012,Johns-Krull2016,Bouma2020}.

@@

@@

\caption{Selected system parameters for TIC\,141146667.}

@@

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.925\textwidth]{figures/sf1.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Evolution of TIC\,141146667 during the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps caused by light from Earth (BTJD 3356-3358.5) and Moon
  (BTJD 3359.5-3360.5) are visible.  Raw two minute
  data are in gray; black time-averages to ten minute sampling.  {\bf
  b-c,} Folded TESS light curve before and after spectroscopy.  Black
  now phase-averages to 100 points per
  cycle.  During BTJD 3352-3356, a state switch occurred near BTJD
  3353, and the dip at $\phi$$\approx$0.8 disappeared.  While the
  large flux decrement was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

We identified TIC\,141146667 in previous work \citep{Bouma2024} by
searching TESS two-minute data from 2018-2022 for highly
structured, periodic light curves \citep{Ricker2015}.  We chose the
star for spectroscopy because its brightness and rapid rotation enabled
an efficient search for variability in its line profiles.
As an apparently single pre-main sequence M dwarf, its properties are
typical for the CPV population (see Table~\ref{tab:params} and
Appendix~\ref{sec:stparams}).

TESS observed TIC\,141146667 ($T$=13.3) for six non-contiguous months
spanning 2019-2024.  TESS acquired these observations at two-minute
cadence during Sectors 41, 48 (TESS DDT039, PI: Kunimoto), and 75
(TESS Program G06030, PI: Bouma).   The 30-minute data during Sectors
14, 15, and 21 probably smeared sharp features over the star's 3.93\,hour
period (see \citealt{Gunther2022}).  The movie in
Figure~\ref{fig:lc} shows the two-minute data: 
like other CPVs, TIC\,141146667 maintains a fixed period over
timescales of years while its detailed photometric morphology evolves.
The nearest known star, TIC\,141146666 ($T$=14.5), is 25$''$ from
TIC\,141146667 and is photometrically stable in the TESS images;
crowding is not a concern.

Although the data gap is unfortunate, Figure~\ref{fig:fulllc} shows that
both before and after the HIRES data were acquired, a large flux
decrement spanned roughly half of each cycle.  From BTJD 3352-3356, this
dip had two sharp local minima;  the minimum at $\phi$$\approx$0.8
decreased in depth following the flare at BTJD 3353, yielding a dip more
closely resembling an asymmetric ``V'' than a ``W''.  Similar CPV state
changes have been previously noted \citep{Stauffer2017,Bouma2024}.  The
photometric shape therefore evolved during the twelve cycles spanning
the 3356-3358 gap, since the average shape from 3358-3363 more closely
resembles the initial ``W''.  Nonetheless, the general photometric
morphology---a small brightening over 45\% of the period, followed by a
complex flux dip spanning 55\% of the period---is similar before and
after the data gap.

Figure~\ref{fig:spec} compiles the TESS and HIRES data from February
2024.  The HIRES spectra show emission in H$\alpha$ beyond the star's
equatorial velocity, $v_{\rm eq}$, of 130\,\kms.  There are at least
two distinct emission components, separated by 180$^\circ$ in phase.
The inner component at lower velocities has clearer sinusoidal
behaviour in time and is double-peaked, with peak semi-amplitudes of
2.07\,$v_{\rm eq}$ and 2.88\,$v_{\rm eq}$, measured following the
procedure in Appendix~\ref{subsec:halpha}.  There is significant
non-periodic variability in the emissivity of this double-peaked
component: the flux excess begins with an amplitude 70\% that of the
continuum, and diminishes to 10\% by the end.  The higher-velocity
component 180$^\circ$ opposite in phase is detected only from
$\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this outer component appears
connected to the star in velocity space.  While its peak
semi-amplitude of 3.88\,$v_{\rm eq}$ is achieved at both $\phi$=0.25
and 0.75, its amplitude similarly decreases from a 60\% excess over
the continuum at the beginning of the observation sequence to a 10\%
excess by its end.  
From the same procedure in Appendix~\ref{subsec:halpha},
we measured
the period for all three spectroscopic
emission components to be consistent with the photometric \periodhr\ hour
period, to within two minutes for the lower-velocity component.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and velocities, this material's motion is
not controlled by gravitational attraction to the star; it is 
a plasma being dragged along with a rigidly rotating stellar magnetic
field.  The velocity semi-amplitude of the sinusoids gives the mean
distance of each clump from the stellar surface: 2.07\,$R_\star$ and
2.88\,$R_\star$ for the inner clumps, and 3.88\,$R_\star$ for the outer
clump.   These clumps transit in front of the star when passing from
negative to positive velocity.  The transits of the two inner clumps
last $\approx$22\% of each cycle, from $\phi$=-0.1 to $\phi$=+0.12 (see
Appendix~\ref{subsec:halpha}).  This spectroscopic transit coincides
with the latter half of the complex eclipse feature in the TESS data.

The H$\alpha$ line core is more complex.  At $|\Delta v / v_{\rm
eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; circumstellar material might also modulate the line
profile.  Figure~\ref{fig:spec}e suggests line core variability caused
by both bright and dark regions on the star's surface, superposed with
smaller-amplitude variability from the transiting circumstellar
material.  For instance, from $\phi$=0-0.3, the double-peaked emission
feature is visible when viewed both on and off-limb; this feature is
circumstellar in origin.  However, the large emission feature that
crosses the star from $\phi$=0.4-0.9 emits at an amplitude greater
than that observed from the circumstellar components, and it crosses
the stellar velocity surface at a speed that suggests it instead
originates from a chromospherically bright region on the star's
surface.  Similarly, from $\phi$=0.6-1.15 a 20\% deep absorption
feature slowly crosses the H$\alpha$ line profile.  This feature
suggests either a chromospherically dark region (e.g.~a starspot
group) crossing the star's surface, or an azimuthally extended
absorptive component to the circumstellar material which is not
visible in emission.  While the origin of the other bright and dark
streaks passing across the line core are similarly ambiguous, a final
exercise to quantify the behavior of the line core is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux correlate with the broadband variability
throughout most of the light curve, except near $\phi$$\approx$0.5,
corresponding to the transit of the 3.9\,$R_\star$ clump and the
occultation of the lower-velocity clump.

@@

\subsection{Physical Properties of the Emitting Region}

% What can be inferred with confidence?

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they demonstrate that corotating circumstellar plasma
clumps exist around at least one such star.  More specifically, the
spectra require plasma with a significant population of hydrogen in the
$n$=3 excited state, with minimal evidence for higher-order excitations
(see Appendix~\ref{subsec:specvar}).  Most of this H$\alpha$ emission
originates in ``clumps'' with size comparable to the star; radial
``spokes'' or azimuthally-extended ``arcs'' for the emitting material
are ruled out by the $\approx$30\,\kms\ H$\alpha$ velocity dispersion
(see Appendix~\ref{subsec:halpha}).  The H$\alpha$ line luminosity
suggests characteristic number densities and masses for the gaseous
component of these clumps of $n_{\rm H} \sim 10^{11}$\,cm$^{-3}$ and
$M_{\rm gas} \sim 10^{17}$\,g (see Appendix~\ref{subsec:gas}).  Dust is
independently constrained to have a total mass $M_{\rm dust} <
10^{17}\,{\rm g}$ based on the lack of a WISE infrared excess; if one
assumes that the opacity in the TESS flux dips comes from dust, a lower
limit $M_{\rm dust} > 10^{15}\,{\rm g}$ follows (see
Appendix~\ref{subsec:dust}).

\subsection{Clumps Within Warped Plasma Torii}

While the H$\alpha$ emission seems to mostly originate from clumps, the
complex flux dip in TIC\,141146667 spans roughly half of each cycle; in
CPVs more broadly, individual dips can last 5-50\% of each cycle, and
they are often distributed in longitude \citep{Bouma2024}.  This raises
the possibility that azimuthally-distributed material may be the norm
for CPVs, and that a warped torus may be a more accurate picture than a
clump.

\citet{Townsend2005} outlined the physics of how rapidly-rotating stars
with strong magnetic fields can support plasma tori.  When the
magnetospheric radii $R_{\rm m}$ of such stars exceeds their Keplerian
corotation radii $R_{\rm c}$, the effective potential experienced by
charged particles has a local minimum outside $R_{\rm c}$, which enables
charged material to accumulate in a torus
\citep{Petit2013,Daley-Yates2024}.  Warps can occur when there is
misalignment between the spin and magnetic axes.  In general, the
material in these centrifugally-supported magnetospheres need neither
transit nor be opaque in broadband optical light.

In a follow-up study, \citet{Townsend2008} synthesized light curves for
these systems, assuming that the optical depth scaled linearly with
density.  They found that W-shaped eclipses, similar to those seen for
TIC\,141146667, can occur when the spin and magnetic axes are moderately
(15-45$^\circ$) misaligned.  The movies associated with their work are
available
online\footnote{\url{http://user.astro.wisc.edu/\~townsend/static.php?ref=rrm-movies\#Download\_Bundles}
last accessed 23 April 2025.}; for a $\omega/\omega_{\rm
c}$$\approx$0.5, edge-on system like TIC\,141146667, the photometric
eclipses have the correct shape, and the H$\alpha$ emission similarly
exhibits double-peaked behavior.  The plasma in this model accumulates
at antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic planes (see Equation~22 of
\citealt{Townsend2005}).

To summarize, the lines of evidence in favor of warped plasma tori for
TIC\,141146667 and CPVs more generally are that {\it i)}
Figure~\ref{fig:spec} shows two emitting plasma clumps separated
180$^\circ$ in phase, with double-peaked H$\alpha$ morphology at
quadrature; {\it ii)} the ``W'' shaped photometric eclipses are
predictions of such a model; {\it iii)} CPV dips last 5-50\% of each
cycle, suggesting an extended azimuthal distribution of circumstellar
material.

\subsection{Origin of CPV Photometric Variability and Astrophysical Analogs}

Microphysically, it is not obvious whether hydrogen alone can produce
the broadband chromatic flux variations seen in CPVs.  The
alternative idea is that charged dust could provide most of the
opacity \citep{Sanderson2023}.  For
TIC\,141146667, Figures~\ref{fig:fulllc}
and~\ref{fig:spec} show that the transits of the H$\alpha$ emitting
material only partially overlap the complex photometric modulation.
This implies that some of the complex photometric dips must involve
additional opacity sources or spatially distinct structures.

Independent of this opacity question, our observations show that
corotating clumps of cool plasma exist around a CPV.  While starspots do
contribute smooth signals to CPV photometric variability, the existence
of these clumps  would not be predicted by a ``starspot-only'' scenario
\citep{Koen2021}.  Scenarios in which the circumstellar material is made
only of dust are similarly ruled out.

The circumstellar material -- either pure plasma or dusty plasma --
could originate either from the star or from an external source.
Plausible external sources include an undetected old disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io torus (e.g.~\citealt{Bagenal1981}),
although with a very different composition.

The other analog for the CPVs are the $\sigma$~Ori~E variables, a rare
subset of B stars with radiatively-driven winds which accumulate into
warped plasma tori \citep{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of these clumps produce
broadband photometric variability through bound-free scattering
\citep{Townsend2005} and Thomson scattering \citep{Berry2022}.  For
$\sigma$~Ori~E and almost all of its analogs, the result is simple
light curves that resemble those of eclipsing binaries, and time-dynamic
H$\alpha$ spectra similar to those in Figure~\ref{fig:spec}.  The two
known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \citep{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong contributions
from higher-order magnetic moments \citep{Kochukhov2011,Shultz2018}.
The main implication is that the photometric complexity of CPVs may be a
direct consequence of magnetic fields with highly multipolar
contributions.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been previously observed to exhibit sinusoidal
time-varying Balmer emission
\citep{Donati2000,Townsend2005,Dunstone2006,Skelly2008}, similar to 
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \citep{Bouma2024}.   One interpretation for the
spectroscopic variability of these stars, and that of the analogous
transient absorption features
\citep{CollierCameron1989,CollierCameron1992,Cang2020}, comes from an
analogy to quiescent solar prominences, cool condensations of plasma in
the solar corona that can last days to weeks \citep{VialEngvold2015}.
In the case of the Sun, these condensations fall back to the photosphere
because gravity is stronger than any magnetic tension or centrifugal
force capable of sustaining them.  However it has been understood at
least since work by \citet{Donati2000} that these ``prominence systems''
can be longitudinally extended, forming a trapped ring of plasma.  While
the issue of why these previously known spectroscopic variables do
not show complex photometric variability remains open, plausible
explanations include that they are not in the required transiting
geometry, or that they lack the necessary source of opacity.  What we
have in the case of TIC\,141146667 is a system that finally exhibits
both sets of phenomena.

@@

Pressing issues include determining the composition and origin of the
circumstellar material, understanding the exact role of the stellar
magnetic field, and exploring the implied space weather experienced by
the close-in rocky exoplanets that, statistically \citep{Dressing2015},
are likely to be present in most CPV systems.

@@

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the modeling perspective,
there is an urgent need for rigid-field magnetohydrodynamic models to
go beyond previous work \citep{Townsend2005,Townsend2008,Krticka2022}
and to explore what field topologies might explain the observed light
curve morphologies.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \citep{Brown2020}; such fields
would yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent constraint on magnetic field topology may also come from
radio observations, which have shown \citep{Kaur2024} that CPVs emit
variable radio signals, including persistent and polarized components.
Detecting radio emission produced by an electron cyclotron maser
instability (e.g.~\citealt{Callingham2021}) in particular would provide
a measurement of the field strength at the site of the emitting region,
further clarifying the magnetosphere's structure.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\citep{Dressing2015}.  However, a few percent of young M dwarfs show
the CPV phenomenon \citep{Rebull2020}, and our data show that some of
the complex photometric features occur when clumps of circumstellar
material transit the star.  The suggested geometric correction implies
that an appreciable minority ($\gtrsim$10\%) of young M dwarfs -- the
rapidly rotating ones with centrifugal magnetospheres -- host
circumstellar environments similar to the CPVs.  Future studies that
combine spectroscopic, polarimetric, and multi-wavelength
observations, along with magnetohydrodynamic modeling, will be key to
understanding the complex magnetospheric environments of these young
stars.

\begin{acknowledgements}
  We thank M.~Jardine, A.~Weinberger, B.~Tofflemire,
  J.~Spake, J.~Winn, and L.~Hillenbrand for insightful conversations
  that significantly informed this work, and to A.~Howard and
  H.~Isaacson for their assistance reducing the HIRES spectra.
